### 2023-09-29 chapters/paper_benchmarking.tex
Updates to citation commands and footnotes

\section{Introduction}
After the introduction of the variational autoencoder (VAE,  \textcite{kingma_autoencoding_2014, rezende_stochastic_2014}) quickly came two temporal extensions for modeling speech data \cite{chung_recurrent_2015, fraccaro_sequential_2016}. Since then, temporal LVMs have undergone little development compared to their counterparts in the image domain, where LVMs recently showed superior performance to autoregressive models such as PixelCNN \cite{oord_pixel_2016, oord_conditional_2016, salimans_pixelcnn_2017}. The improvements in the image domain have been driven mainly by top-down inference models and deeper latent hierarchies \cite{sonderby_ladder_2016, maaloe_biva_2019, vahdat_nvae_2020, child_very_2021,sinha_consistency_2021,kingma_variational_2021}. In speech modeling however, autoregressive models such as the WaveNet remain state-of-the-art \cite{oord_wavenet_2016}.

To compare and develop LVMs for speech, we need good benchmarks similar to those in the image domain. Image benchmarks commonly compare likelihood scores, but research in the speech domain often omits reporting a likelihood \cite{oord_wavenet_2016, hsu_unsupervised_2017, oord_neural_2018} or report likelihoods that are incomparable due to subtle differences in the assumed data distribution \cite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_unsupervised_2017, aksan_stcn_2019}. Without a proper standard, it is difficult to compare explicit likelihood models for speech and develop them in an informed manner.

To advance the state of LVMs for speech, this paper (i) develops a benchmark for LVMs based on model likelihood, (ii) introduces a hierarchical LVM architecture without autoregressive decoder, (iii) compares LVMs to deterministic counterparts including WaveNet, and (iv) qualitatively and quantitatively evaluates the latent variables learned by different LVMs based on their usefulness for phoneme recognition. We find that:
\begin{itemize}[leftmargin=8mm]
    \setlength\itemsep{-0em}
    \item [(I)] State-of-the-art LVMs achieve likelihoods that are inferior to WaveNet at high temporal resolution but are superior at lower resolutions. Interestingly, we find that a standard LSTM \cite{hochreiter_long_1997} almost matches the likelihood of WaveNet.
    \item [(II)] LVMs with powerful autoregressive decoders achieve better likelihoods than the non-autoregressive LVM.
    \item[(III)] The expressiveness of LVMs for speech increases with a deeper hierarchy of stochastic latent variables, similar to conclusions within image modeling. % \cite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
    \item[(IV)] LVMs learn rich representations that are as good or better than Mel spectrograms for phoneme recognition also when using only 10 minutes of labeled data.
\end{itemize}
At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models
\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}.
% \footnote{\fontsize{8.5}{0}\selectfont\texttt{ \url{github.com/anonymous/available-later}}}.
Before presenting the results, we provide a brief survey of existing LVMs for speech in a coherent notation.

@@

% \linepenalty=1000
% \subsection{Related work}
LVMs formulated as VAEs continue to be of interest since they are able to learn an approximation to the posterior distribution of assumed latent variables. The posterior is usually of a reduced dimensionality compared to the input and lies close to a known prior distribution. Approximate posteriors have various applications e.g. semi-supervised learning \cite{kingma_semi-supervised_2014} and anomaly detection \cite{havtorn_hierarchical_2021}.

In recent years, several complementary methods have been proposed to improve the expressiveness of VAEs. These include building more expressive priors via methods such as normalizing flows \cite{rezende_variational_2015} and building a deeper hierarchy of stochastic latent variables such as the Ladder VAE \cite{sonderby_ladder_2016}. In this research, we focus on the latter due to the recent breakthroughs in image modeling using VAEs without costly autoregressive dependencies on the observed variable \cite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.

Several works have applied LVMs to speech. Among the first contributions were the VRNN \cite{chung_recurrent_2015} and SRNN \cite{fraccaro_sequential_2016} which can be seen as conditional VAEs per timestep. Other recent LVMs include the FH-VAE \cite{hsu_unsupervised_2017}, which leverages an additional latent variable to capture global features, and Z-forcing \cite{goyal_zforcing_2017}, which resembles the SRNN but includes an auxiliary task in the latent space to increase its utilization. The VQ-VAE \cite{oord_neural_2018} is a hybrid between an LVM and an autoregressive model which uses a quantized latent space to improve the quality of generated samples. The Stochastic WaveNet \cite{lai_stochastic_2018} and STCN \cite{aksan_stcn_2019} use WaveNet encoder and decoders and temporally independent latent variables.

In this paper, we focus on the VRNN, SRNN and STCN. We exclude the Stochastic WaveNet as it is similar to STCN and achieves inferior likelihoods \cite{aksan_stcn_2019}. The FH-VAE, with disjoint latent variables and discriminative objective, Z-forcing, with an auxiliary task, and the VQ-VAE, with a quantized latent space and autoregressive prior fitted after training, all introduce significant changes to the original VAE framework and are also not included here.

All selected models have autoregressive generative models which let future observed variables be generated by conditioning on previously generated values. Inspired by recent progress in the image domain, we therefore formulate and benchmark a novel temporal LVM which does not rely on an autoregressive decoder. We do so by adapting the hierarchical Clockwork Variational Autoencoder \cite{saxena_clockwork_2021}, originally proposed for video generation, to speech.

\subsection{Sequential deep latent variable models}
The selected models are all sequential deep latent variable models trained with variational inference and the reparameterization trick \cite{kingma_autoencoding_2014}.
% The VRNN, SRNN, STCN and CW-VAE are all autoencoders trained with variational inference and the reparameterization trick \cite{kingma_autoencoding_2014}. 
They take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
% The VRNN, SRNN, STCN and CW-VAE are all VAEs \cite{kingma_autoencoding_2014} that take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
We let $\xb_{1:T}$ refer to the observed variable or a downsampled version of it. We will sometimes use $\xb$ to refer to the sequence $\xb_{1:T}$ when it is not ambiguous.
% In general, $\xb_{1:T}$ may refer to the original observed variable or a deterministically downsampled representation of the observed variable. We will sometimes use $\xb$ to refer to the full sequence $\xb_{1:T}$ when it is not ambiguous.

@@

\subsection{Variational recurrent neural network ({\small VRNN})}
The VRNN \cite{chung_recurrent_2015} is essentially a VAE per timestep. At timestep $t$, the VAE is conditioned on the hidden state of a recurrent neural network (RNN), $\db_{t}\in\mathbb{R}^{D_d}$, with state transition $\db_{t} = f([\xb_{t-1}, \zb_{t-1}], \db_{t-1})$ where $[\cdot, \cdot]$ denotes concatenation. 
The VRNN uses a Gated Recurrent Unit (GRU, \textcite{cho_properties_2014}) for $f$.
The joint distribution factorizes over time and the latent variables are autoregressive in both the observed and latent space,
\begin{equation}
    p(\xb_{1:T},\zb_{1:T}) = \prod_{t=1}^{T} p(\xb_t|\xb_{<t},\zb_{\leq t}) p(\zb_t|\xb_{<t},\zb_{<t}) \enspace . \label{eq: vrnn joint factorization}
\end{equation}
The approximate posterior similarly factorizes over time,
\begin{equation}
    q(\zb_{1:T}|\xb_{1:T}) = \prod_{t=1}^{T} q(\zb_t|\xb_{\leq t},\zb_{<t}) \enspace . \label{eq: vrnn inference factorization}
\end{equation}
From this, the ELBO for the VRNN is
\begin{equation}
    \mathcal{L}(\xb) = ~ \mathbb{E}_{q(\zb_{1:T}|\xb_{1:T})} \bigg[ \sum_{t} \log p(\xb_t|\xb_{<t},\zb_{\leq t}) - D_\text{KL}\left( q(\zb_t|\xb_{\leq t},\zb_{<t}) \parallel p(\zb_t|\xb_{<t},\zb_{<t}) \right)  \bigg] \label{eq: vrnn elbo} \enspace .
\end{equation}
% \begin{equation}
% \begin{split}
%     \mathcal{L}(\xb) =&~ \mathbb{E}_{q(\zb_{1:T}|\xb_{1:T})} \bigg[ \sum_{t} \log p(\xb_t|\xb_{<t},\zb_{\leq t}) \\ & - \text{KL}\left( q(\zb_t|\xb_{\leq t},\zb_{<t}) \parallel p(\zb_t|\xb_{<t},\zb_{<t}) \right)  \bigg] \enspace .
%     \label{eq: vrnn elbo}
% \end{split}
% \end{equation}
The VRNN uses diagonal covariance Gaussian distributions $\mathcal{N}$ for the prior and posterior distributions. We denote the output distribution of choice by $\mathcal{D}$. 
% \begin{align}
%     q(\zb_t|\xb_{\leq t},\zb_{<t}) &= \mathcal{N}\left(\alphab_{q}(\xb_{t},\db_{t})\right) \label{eq: vrnn parameterization inference} \\
%     p(\zb_t|\xb_{<t},\zb_{<t}) &= \mathcal{N}\left(\alphab_{p}(\db_t)\right) \label{eq: vrnn parameterization prior} \\
%     p(\xb_t|\xb_{<t},\zb_{\leq t}) &= \mathcal{D}\left(\betab(\zb_{t},\db_{t})\right) \enspace . \label{eq: vrnn parameterization observation}
% \end{align}
\begin{gather}
    p(\zb_t|\xb_{<t},\zb_{<t}) = \mathcal{N}\left(\alphab_{p}(\db_t)\right) \enspace , \nonumber \\
    p(\xb_t|\xb_{<t},\zb_{\leq t}) = \mathcal{D}\left(\betab(\zb_{t},\db_{t})\right) \enspace , \label{eq: vrnn parameterization generative} \\
    q(\zb_t|\xb_{\leq t},\zb_{<t}) = \mathcal{N}\left(\alphab_{q}(\xb_{t},\db_{t})\right) \enspace . \nonumber 
    \label{eq: vrnn parameterization inference}
\end{gather}
All sets of distributional parameters, $\alphab_q,\alphab_p$ and $\betab$, are the outputs of densely connected neural networks which we notationally overload as functions in equations~\cref{eq: vrnn parameterization generative} and \cref{eq: vrnn parameterization inference}. It is common to refer to $\alphab_q$ as the inference model or encoder and $\betab$ as the decoder. Together with $\betab$, the structural model $\alphab_p$ forms the generative model.

Since the decoder is dependent on $\db_t$, the transition function $f$ allows the VRNN to learn to ignore parts of or the entire latent variable and establish a purely deterministic transition from $\xb_{t-1}$ to $\db_t$ (\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}). This failure mode is commonly referred to as posterior collapse and is a well-known weakness of VAEs with powerful decoders \cite{bowman_generating_2016, sonderby_ladder_2016}.

\subsection{Stochastic recurrent neural network ({\small SRNN})}
The SRNN \cite{fraccaro_sequential_2016} is similar to the VRNN but differs by separating the stochastic latent variables from the deterministic representations (\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}). That is, the GRU state transition is independent of $\zb_{1:T}$ such that $\db_{t} = f(\xb_{t-1}, \db_{t-1})$. With this, the joint $p(\xb_{1:T},\zb_{1:T})$ can be written as for the VRNN in \cref{eq: vrnn joint factorization}. The approximate posterior of $\zb_t$ is conditioned on the full observed sequence,
\begin{align}
    \label{eq: srnn inference factorization}
    q(\zb_{1:T}|\xb_{1:T}) = \prod_{t=1}^{T} q(\zb_t|\xb_{1:T},\zb_{t-1}) \enspace .
\end{align}
This is achieved by introducing a second GRU that runs backwards in time with transition $\ab_t = g([\xb_t,\db_t], \ab_{t+1})$. While $p(\xb_t|\xb_{<t},\zb_{\leq t})$ remains as in \cref{eq: vrnn parameterization generative}, we have
\begin{gather}
        p(\zb_t|\xb_{<t},\zb_{<t}) = \mathcal{N}\left(\alphab_{p}(\zb_{t-1},\db_t)\right) \enspace , \nonumber \\
        q(\zb_t|\xb_{1:T},\zb_{<t}) = \mathcal{N}\left(\alphab_{q}(\zb_{t-1},\ab_{t})\right) \enspace . \label{eq: srnn parameterization}
    % p(\xb_t|\xb_{<t},\zb_{\leq t}) &= \mathcal{D}\left(\vrho(\zb_{t},\db_{t})\right) \enspace . \label{eq: srnn parameterization observation}
\end{gather}
By inferring $\zb_t$ conditioned on the full sequence $\xb_{1:T}$, the SRNN performs smoothing. This has been noted to better approximate the true posterior of $\zb_t$ which can be shown to depend on the full observed sequence \cite{bayer_mind_2021}. Comparatively, the VRNN performs filtering.

@@

\subsection{Stochastic temporal convolutional network ({\small STCN})}
The STCN \cite{aksan_stcn_2019} is a hierarchical latent variable model with an autoregressive generative model based on WaveNet \cite{oord_wavenet_2016}. Contrary to VRNN and SRNN, the latent variables have no transition functions connecting them over time.
Instead, a latent $\zb_t^{(l)}$ at layer $l$ is conditioned on a window of observed variables $\xb_{\scriptscriptstyle\mathcal{R}_t^{(l)}}$ via a WaveNet encoder where $\mathcal{R}_t^{(l)} = \cbra{t-r_l+1,\dots,t}$ and $r_l$ is the receptive field of the encoder at layer $l$. The window size grows exponentially with $l$.
\begin{equation}
    p(\xb_{1:T},\zb_{1:T}^{(1:L)}) = \prod_{t=1}^{T} p(\xb_t|\zb^{(1:L)}_{{\scriptscriptstyle \mathcal{R}}_{t}^{(1)}}) \prod_{l=1}^{L} p(\zb^{(l)}_{t}|\xb_{\scriptscriptstyle\mathcal{R}_{t-1}^{(l)}},\zb^{(l+1)}_{t}) \enspace , \nonumber
    \label{eq: stcn joint factorization}
\end{equation}
where $\zb_t^{(L+1)}:=\emptyset$ for notational convenience. 
The deterministic encoding is $\db_t^{(l)} = h(\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}})$ where $h$ is the encoder and $\db_t^{(l)}$ is extracted from the $l$'th layer similar to a Ladder VAE \cite{sonderby_ladder_2016}. 
The approximate posterior is
\begin{equation}
    q\pa{\zb_{1:T}^{(1:L)}|\xb_{1:T}} = \prod_{t=1}^{T}\prod_{l=1}^{L} q\pa{\zb_{t}^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}},\zb_t^{(l+1)}} \enspace .
\end{equation}
The factorized distributions are given as
% \begin{align}
%     q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}},\zb_t^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{q}^{(l)}\pa{\zb_t^{(l+1)},\db_t^{(l)}}} \enspace , \\
%     p\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t-1}^{(l)}},\zb_{t}^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{p}^{(l)}\pa{\zb_t^{(l+1)},\db_{t-1}^{(l)}}} \enspace \nonumber , \\
%     p\left(\xb_t|\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}\right) &= \mathcal{D}\pa{\betab\pa{\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}}} \enspace . \nonumber \label{eq: stcn parameterization} \\
% \end{align}
\begin{gather}
    p\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t-1}^{(l)}},\zb_{t}^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{p}^{(l)}\pa{\zb_t^{(l+1)},\db_{t-1}^{(l)}}} \, \enspace , \nonumber \\
    p\left(\xb_t|\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}\right) = \mathcal{D}\pa{\betab\pa{\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}}} \enspace , \label{eq: stcn parameterization} \\
    q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}},\zb_t^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{q}^{(l)}\pa{\zb_t^{(l+1)},\db_t^{(l)}}} \enspace . \nonumber
\end{gather}
The decoder $\betab\left(\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}\right)$ is also a WaveNet.

>>>>

\section{Introduction}
After the introduction of the variational autoencoder (VAE,  \textcite{kingma_autoencoding_2014, rezende_stochastic_2014}) quickly came two temporal extensions for modeling speech data \parencite{chung_recurrent_2015, fraccaro_sequential_2016}. Since then, temporal LVMs have undergone little development compared to their counterparts in the image domain, where LVMs recently showed superior performance to autoregressive models such as PixelCNN \parencite{oord_pixel_2016, oord_conditional_2016, salimans_pixelcnn_2017}. The improvements in the image domain have been driven mainly by top-down inference models and deeper latent hierarchies \parencite{sonderby_ladder_2016, maaloe_biva_2019, vahdat_nvae_2020, child_very_2021,sinha_consistency_2021,kingma_variational_2021}. In speech modeling however, autoregressive models such as the WaveNet remain state-of-the-art \parencite{oord_wavenet_2016}.

To compare and develop LVMs for speech, we need good benchmarks similar to those in the image domain. Image benchmarks commonly compare likelihood scores, but research in the speech domain often omits reporting a likelihood \parencite{oord_wavenet_2016, hsu_unsupervised_2017, oord_neural_2018} or report likelihoods that are incomparable due to subtle differences in the assumed data distribution \parencite{chung_recurrent_2015, fraccaro_sequential_2016, hsu_unsupervised_2017, aksan_stcn_2019}. Without a proper standard, it is difficult to compare explicit likelihood models for speech and develop them in an informed manner.

To advance the state of LVMs for speech, this paper (i) develops a benchmark for LVMs based on model likelihood, (ii) introduces a hierarchical LVM architecture without autoregressive decoder, (iii) compares LVMs to deterministic counterparts including WaveNet, and (iv) qualitatively and quantitatively evaluates the latent variables learned by different LVMs based on their usefulness for phoneme recognition. We find that:
\begin{itemize}[leftmargin=8mm]
    \setlength\itemsep{-0em}
    \item [(I)] State-of-the-art LVMs achieve likelihoods that are inferior to WaveNet at high temporal resolution but are superior at lower resolutions. Interestingly, we find that a standard LSTM \parencite{hochreiter_long_1997} almost matches the likelihood of WaveNet.
    \item [(II)] LVMs with powerful autoregressive decoders achieve better likelihoods than the non-autoregressive LVM.
    \item[(III)] The expressiveness of LVMs for speech increases with a deeper hierarchy of stochastic latent variables, similar to conclusions within image modeling. % \parencite{maaloe_biva_2019, vahdat_nvae_2020, child_very_2021}.
    \item[(IV)] LVMs learn rich representations that are as good or better than Mel spectrograms for phoneme recognition also when using only 10 minutes of labeled data.
\end{itemize}
At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models
\footnote{\href{https://github.com/JakobHavtorn/benchmarking-lvms}{\url{github.com/JakobHavtorn/benchmarking-lvms}}}.
% \footnote{\fontsize{8.5}{0}\selectfont\texttt{ \url{github.com/anonymous/available-later}}}.
Before presenting the results, we provide a brief survey of existing LVMs for speech in a coherent notation.

@@

% \linepenalty=1000
% \subsection{Related work}
LVMs formulated as VAEs continue to be of interest since they are able to learn an approximation to the posterior distribution of assumed latent variables. The posterior is usually of a reduced dimensionality compared to the input and lies close to a known prior distribution. Approximate posteriors have various applications e.g. semi-supervised learning \parencite{kingma_semi-supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}.

In recent years, several complementary methods have been proposed to improve the expressiveness of VAEs. These include building more expressive priors via methods such as normalizing flows \parencite{rezende_variational_2015} and building a deeper hierarchy of stochastic latent variables such as the Ladder VAE \parencite{sonderby_ladder_2016}. In this research, we focus on the latter due to the recent breakthroughs in image modeling using VAEs without costly autoregressive dependencies on the observed variable \parencite{maaloe_biva_2019,vahdat_nvae_2020,child_very_2021}.

Several works have applied LVMs to speech. Among the first contributions were the VRNN \parencite{chung_recurrent_2015} and SRNN \parencite{fraccaro_sequential_2016} which can be seen as conditional VAEs per timestep. Other recent LVMs include the FH-VAE \parencite{hsu_unsupervised_2017}, which leverages an additional latent variable to capture global features, and Z-forcing \parencite{goyal_zforcing_2017}, which resembles the SRNN but includes an auxiliary task in the latent space to increase its utilization. The VQ-VAE \parencite{oord_neural_2018} is a hybrid between an LVM and an autoregressive model which uses a quantized latent space to improve the quality of generated samples. The Stochastic WaveNet \parencite{lai_stochastic_2018} and STCN \parencite{aksan_stcn_2019} use WaveNet encoder and decoders and temporally independent latent variables.

In this paper, we focus on the VRNN, SRNN and STCN. We exclude the Stochastic WaveNet as it is similar to STCN and achieves inferior likelihoods \parencite{aksan_stcn_2019}. The FH-VAE, with disjoint latent variables and discriminative objective, Z-forcing, with an auxiliary task, and the VQ-VAE, with a quantized latent space and autoregressive prior fitted after training, all introduce significant changes to the original VAE framework and are also not included here.

All selected models have autoregressive generative models which let future observed variables be generated by conditioning on previously generated values. Inspired by recent progress in the image domain, we therefore formulate and benchmark a novel temporal LVM which does not rely on an autoregressive decoder. We do so by adapting the hierarchical Clockwork Variational Autoencoder \parencite{saxena_clockwork_2021}, originally proposed for video generation, to speech.

\subsection{Sequential deep latent variable models}
The selected models are all sequential deep latent variable models trained with variational inference and the reparameterization trick \parencite{kingma_autoencoding_2014}.
% The VRNN, SRNN, STCN and CW-VAE are all autoencoders trained with variational inference and the reparameterization trick \parencite{kingma_autoencoding_2014}. 
They take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
% The VRNN, SRNN, STCN and CW-VAE are all VAEs \parencite{kingma_autoencoding_2014} that take as input a variable-length sequence $\xb_{1:T} = \left(\xb_1, \xb_2, \dots, \xb_{T}\right)$ with $\xb_t \in \mathbb{R}^{D_x}$. 
We let $\xb_{1:T}$ refer to the observed variable or a downsampled version of it. We will sometimes use $\xb$ to refer to the sequence $\xb_{1:T}$ when it is not ambiguous.
% In general, $\xb_{1:T}$ may refer to the original observed variable or a deterministically downsampled representation of the observed variable. We will sometimes use $\xb$ to refer to the full sequence $\xb_{1:T}$ when it is not ambiguous.

@@

\subsection{Variational recurrent neural network ({\small VRNN})}
The VRNN \parencite{chung_recurrent_2015} is essentially a VAE per timestep. At timestep $t$, the VAE is conditioned on the hidden state of a recurrent neural network (RNN), $\db_{t}\in\mathbb{R}^{D_d}$, with state transition $\db_{t} = f([\xb_{t-1}, \zb_{t-1}], \db_{t-1})$ where $[\cdot, \cdot]$ denotes concatenation. 
The VRNN uses a Gated Recurrent Unit (GRU, \textcite{cho_properties_2014}) for $f$.
The joint distribution factorizes over time and the latent variables are autoregressive in both the observed and latent space,
\begin{equation}
    p(\xb_{1:T},\zb_{1:T}) = \prod_{t=1}^{T} p(\xb_t|\xb_{<t},\zb_{\leq t}) p(\zb_t|\xb_{<t},\zb_{<t}) \enspace . \label{eq: vrnn joint factorization}
\end{equation}
The approximate posterior similarly factorizes over time,
\begin{equation}
    q(\zb_{1:T}|\xb_{1:T}) = \prod_{t=1}^{T} q(\zb_t|\xb_{\leq t},\zb_{<t}) \enspace . \label{eq: vrnn inference factorization}
\end{equation}
From this, the ELBO for the VRNN is
\begin{equation}
    \mathcal{L}(\xb) = ~ \mathbb{E}_{q(\zb_{1:T}|\xb_{1:T})} \bigg[ \sum_{t} \log p(\xb_t|\xb_{<t},\zb_{\leq t}) - D_\text{KL}\left( q(\zb_t|\xb_{\leq t},\zb_{<t}) \parallel p(\zb_t|\xb_{<t},\zb_{<t}) \right)  \bigg] \label{eq: vrnn elbo} \enspace .
\end{equation}
% \begin{equation}
% \begin{split}
%     \mathcal{L}(\xb) =&~ \mathbb{E}_{q(\zb_{1:T}|\xb_{1:T})} \bigg[ \sum_{t} \log p(\xb_t|\xb_{<t},\zb_{\leq t}) \\ & - \text{KL}\left( q(\zb_t|\xb_{\leq t},\zb_{<t}) \parallel p(\zb_t|\xb_{<t},\zb_{<t}) \right)  \bigg] \enspace .
%     \label{eq: vrnn elbo}
% \end{split}
% \end{equation}
The VRNN uses diagonal covariance Gaussian distributions $\mathcal{N}$ for the prior and posterior distributions. We denote the output distribution of choice by $\mathcal{D}$. 
% \begin{align}
%     q(\zb_t|\xb_{\leq t},\zb_{<t}) &= \mathcal{N}\left(\alphab_{q}(\xb_{t},\db_{t})\right) \label{eq: vrnn parameterization inference} \\
%     p(\zb_t|\xb_{<t},\zb_{<t}) &= \mathcal{N}\left(\alphab_{p}(\db_t)\right) \label{eq: vrnn parameterization prior} \\
%     p(\xb_t|\xb_{<t},\zb_{\leq t}) &= \mathcal{D}\left(\betab(\zb_{t},\db_{t})\right) \enspace . \label{eq: vrnn parameterization observation}
% \end{align}
\begin{gather}
    p(\zb_t|\xb_{<t},\zb_{<t}) = \mathcal{N}\left(\alphab_{p}(\db_t)\right) \enspace , \nonumber \\
    p(\xb_t|\xb_{<t},\zb_{\leq t}) = \mathcal{D}\left(\betab(\zb_{t},\db_{t})\right) \enspace , \label{eq: vrnn parameterization generative} \\
    q(\zb_t|\xb_{\leq t},\zb_{<t}) = \mathcal{N}\left(\alphab_{q}(\xb_{t},\db_{t})\right) \enspace . \nonumber 
    \label{eq: vrnn parameterization inference}
\end{gather}
All sets of distributional parameters, $\alphab_q,\alphab_p$ and $\betab$, are the outputs of densely connected neural networks which we notationally overload as functions in equations~\cref{eq: vrnn parameterization generative} and \cref{eq: vrnn parameterization inference}. It is common to refer to $\alphab_q$ as the inference model or encoder and $\betab$ as the decoder. Together with $\betab$, the structural model $\alphab_p$ forms the generative model.

Since the decoder is dependent on $\db_t$, the transition function $f$ allows the VRNN to learn to ignore parts of or the entire latent variable and establish a purely deterministic transition from $\xb_{t-1}$ to $\db_t$ (\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}). This failure mode is commonly referred to as posterior collapse and is a well-known weakness of VAEs with powerful decoders \parencite{bowman_generating_2016, sonderby_ladder_2016}.

\subsection{Stochastic recurrent neural network ({\small SRNN})}
The SRNN \parencite{fraccaro_sequential_2016} is similar to the VRNN but differs by separating the stochastic latent variables from the deterministic representations (\cref{fig: autoregressive vrnn srnn cwvae 1L graphs}). That is, the GRU state transition is independent of $\zb_{1:T}$ such that $\db_{t} = f(\xb_{t-1}, \db_{t-1})$. With this, the joint $p(\xb_{1:T},\zb_{1:T})$ can be written as for the VRNN in \cref{eq: vrnn joint factorization}. The approximate posterior of $\zb_t$ is conditioned on the full observed sequence,
\begin{align}
    \label{eq: srnn inference factorization}
    q(\zb_{1:T}|\xb_{1:T}) = \prod_{t=1}^{T} q(\zb_t|\xb_{1:T},\zb_{t-1}) \enspace .
\end{align}
This is achieved by introducing a second GRU that runs backwards in time with transition $\ab_t = g([\xb_t,\db_t], \ab_{t+1})$. While $p(\xb_t|\xb_{<t},\zb_{\leq t})$ remains as in \cref{eq: vrnn parameterization generative}, we have
\begin{gather}
        p(\zb_t|\xb_{<t},\zb_{<t}) = \mathcal{N}\left(\alphab_{p}(\zb_{t-1},\db_t)\right) \enspace , \nonumber \\
        q(\zb_t|\xb_{1:T},\zb_{<t}) = \mathcal{N}\left(\alphab_{q}(\zb_{t-1},\ab_{t})\right) \enspace . \label{eq: srnn parameterization}
    % p(\xb_t|\xb_{<t},\zb_{\leq t}) &= \mathcal{D}\left(\vrho(\zb_{t},\db_{t})\right) \enspace . \label{eq: srnn parameterization observation}
\end{gather}
By inferring $\zb_t$ conditioned on the full sequence $\xb_{1:T}$, the SRNN performs smoothing. This has been noted to better approximate the true posterior of $\zb_t$ which can be shown to depend on the full observed sequence \parencite{bayer_mind_2021}. Comparatively, the VRNN performs filtering.

@@

\subsection{Stochastic temporal convolutional network ({\small STCN})}
The STCN \parencite{aksan_stcn_2019} is a hierarchical latent variable model with an autoregressive generative model based on WaveNet \parencite{oord_wavenet_2016}. Contrary to VRNN and SRNN, the latent variables have no transition functions connecting them over time.
Instead, a latent $\zb_t^{(l)}$ at layer $l$ is conditioned on a window of observed variables $\xb_{\scriptscriptstyle\mathcal{R}_t^{(l)}}$ via a WaveNet encoder where $\mathcal{R}_t^{(l)} = \cbra{t-r_l+1,\dots,t}$ and $r_l$ is the receptive field of the encoder at layer $l$. The window size grows exponentially with $l$.
\begin{equation}
    p(\xb_{1:T},\zb_{1:T}^{(1:L)}) = \prod_{t=1}^{T} p(\xb_t|\zb^{(1:L)}_{{\scriptscriptstyle \mathcal{R}}_{t}^{(1)}}) \prod_{l=1}^{L} p(\zb^{(l)}_{t}|\xb_{\scriptscriptstyle\mathcal{R}_{t-1}^{(l)}},\zb^{(l+1)}_{t}) \enspace , \nonumber
    \label{eq: stcn joint factorization}
\end{equation}
where $\zb_t^{(L+1)}:=\emptyset$ for notational convenience. 
The deterministic encoding is $\db_t^{(l)} = h(\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}})$ where $h$ is the encoder and $\db_t^{(l)}$ is extracted from the $l$'th layer similar to a Ladder VAE \parencite{sonderby_ladder_2016}. 
The approximate posterior is
\begin{equation}
    q\pa{\zb_{1:T}^{(1:L)}|\xb_{1:T}} = \prod_{t=1}^{T}\prod_{l=1}^{L} q\pa{\zb_{t}^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}},\zb_t^{(l+1)}} \enspace .
\end{equation}
The factorized distributions are given as
% \begin{align}
%     q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}},\zb_t^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{q}^{(l)}\pa{\zb_t^{(l+1)},\db_t^{(l)}}} \enspace , \\
%     p\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t-1}^{(l)}},\zb_{t}^{(l+1)}\right) &= \mathcal{N}\pa{\alphab_{p}^{(l)}\pa{\zb_t^{(l+1)},\db_{t-1}^{(l)}}} \enspace \nonumber , \\
%     p\left(\xb_t|\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}\right) &= \mathcal{D}\pa{\betab\pa{\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}}} \enspace . \nonumber \label{eq: stcn parameterization} \\
% \end{align}
\begin{gather}
    p\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t-1}^{(l)}},\zb_{t}^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{p}^{(l)}\pa{\zb_t^{(l+1)},\db_{t-1}^{(l)}}} \, \enspace , \nonumber \\
    p\left(\xb_t|\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}\right) = \mathcal{D}\pa{\betab\pa{\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}}} \enspace , \label{eq: stcn parameterization} \\
    q\left(\zb_t^{(l)}|\xb_{\scriptscriptstyle\mathcal{R}_{t}^{(l)}},\zb_t^{(l+1)}\right) = \mathcal{N}\pa{\alphab_{q}^{(l)}\pa{\zb_t^{(l+1)},\db_t^{(l)}}} \enspace . \nonumber
\end{gather}
The decoder $\betab\left(\zb^{(1:L)}_{\scriptscriptstyle\mathcal{R}_{t}^{(1)}}\right)$ is also a WaveNet.
